---
Cultivar x Treatment interaction
---

This code performs the cultivar x treatment analysis
This code requires files generated from the following scripts:
OTU_filtering.Rmd
vsd_normalization.Rmd

Load the libraries
```{r}
library(DESeq2)
library(biobroom)
library(dplyr)
library(tidyr)
```

Load the data. Because of the high variability between compartments and soils, we evaluated the cultivar x treatment interactions within subsets covering each compartment-soil combination
```{r}
full.otu <- readRDS("../Data/otu_filt.RDS")
full.map <- readRDS("../Data/map.RDS")
full.map <- mutate(full.map, Group = paste(Cultivar, Treatment, sep = "."))

RS.A.map <- full.map %>% filter(Compartment == "RS" & Soil == "A") 
ES.A.map <- full.map %>% filter(Compartment == "ES" & Soil == "A")
RS.B.map <- full.map %>% filter(Compartment == "RS" & Soil == "B") 
ES.B.map <- full.map %>% filter(Compartment == "ES" & Soil == "B")
RS.D.map <- full.map %>% filter(Compartment == "RS" & Soil == "D") 
ES.D.map <- full.map %>% filter(Compartment == "ES" & Soil == "D")

sets <- c("RS.A", "ES.A","RS.B", "ES.B", "RS.D", "ES.D")

sets.list <- list(RS.A = RS.A.map,
                  ES.A = ES.A.map,
                  RS.B = RS.B.map,
                  ES.B = ES.B.map,
                  RS.D = RS.D.map,
                  ES.D = ES.D.map) 

eff.list <- vector("list")
fc.list <- vector("list")
```

Detect all the interactions using an lrt approach
```{r}
for(s in sets){
  map <- sets.list[[s]]
  row.names(map) <- map$SampleID
  
  otu <- full.otu[,match(map$SampleID, colnames(full.otu))]
  otu <- otu[rowSums(otu) > 0,]

  # Generate DESeq object
  dds <- DESeqDataSetFromMatrix(countData = otu,
                                colData = map,
                                design = ~ Treatment + Cultivar + Treatment:Cultivar + Library)
  
  dds <- DESeq(dds, test = "LRT", reduced = ~ Treatment + Cultivar + Library)
  
  eff.list[[s]] <- results(dds) %>% tidy()
}
```

Format the results
```{r}
eff.df <- plyr::ldply(eff.list, function(x) x)
names(eff.df)[1:2] <- c("Set", "OTU_ID") 
eff.df <- eff.df %>%
  separate(Set, c("Compartment", "Soil"), sep = "\\.") 
```

Calculate the drought effect within compartmnet to obtain the logFC
```{r}
for(s in sets){
  map <- sets.list[[s]]
  row.names(map) <- map$SampleID
  
  otu <- full.otu[,match(map$SampleID, colnames(full.otu))]
  otu <- otu[rowSums(otu) > 0,]

  # Generate DESeq object
  dds <- DESeqDataSetFromMatrix(countData = otu,
                                colData = map,
                                design = ~ Group + Library)
  
  dds <- DESeq(dds)
  
  contrasts = c("G1", "G2", "S1", "S2") 
  
  contrast.list <- list(G1 = c("Group", "G1.DS", "G1.WC"),
                        G2 = c("Group", "G2.DS", "G2.WC"),
                        S1 = c("Group", "S1.DS", "S1.WC"),
                        S2 = c("Group", "S2.DS", "S2.WC"))
  
  # We are using the Bonferroni correction for dealing with multiple comparisons
  
  for(i in contrasts) {
    fc.list[[paste(s,i,sep = ".")]] <- tidy(results(dds, contrast = contrast.list[[i]]))
  }
}
```

Format the results
```{r}
fc.df <- plyr::ldply(fc.list, function(x) x)
names(fc.df)[1:2] <- c("Contrast", "OTU_ID") 
fc.df <- fc.df %>%
  separate(Contrast, c("Compartment", "Soil", "Cultivar"), sep = "\\.") 
```

Identify how many interactions were detected
```{r}
eff.df %>%
  filter(p.adjusted < 0.05) %>%
  group_by(Soil, Compartment) %>%
  count()
```

Generate the plots
```{r}
library(ggplot2)
library(scales)
library(RColorBrewer)

clt.lrt.sig <- eff.df %>% filter(p.adjusted < 0.05)
clt.fc.sig <- fc.df %>%
  inner_join(select(clt.lrt.sig, Soil, Compartment, OTU_ID), by = c("Soil", "Compartment", "OTU_ID"))

otu.vsd <- readRDS("../Data/otu_vsd_blind.RDS")
otu.vsd[otu.vsd < 0] <- 0
otu.vsd.tidy <- data.frame(otu.vsd) %>%
  mutate(OTU_ID = row.names(otu.vsd)) %>%
  gather(key = "SampleID", value = "RelAb", -OTU_ID)
otu.vsd.tidy <- inner_join(otu.vsd.tidy, full.map, by = "SampleID")

otu.vsd.tidy$Treatment = factor(otu.vsd.tidy$Treatment,
                               levels = c("DS", "WC"),
                               labels = c("Drought", "Control"))

maxval  <- ceiling(max(abs(clt.fc.sig$estimate)))
values  <- seq(-maxval,maxval,length = 8)

otu.d <- otu.vsd.tidy %>% 
  inner_join(clt.fc.sig, by = c("Compartment", "Soil", "Cultivar", "OTU_ID")) %>% 
  filter(Soil == "D") %>%
  ggplot(aes(Cultivar, RelAb, fill = Treatment)) +
  geom_boxplot() +
  geom_point(aes(Cultivar, -0.5, color = estimate, alpha = p.adjusted < 0.05), size = 5) +
  facet_wrap(~OTU_ID, scales = "free_y", ncol = 6) +
  scale_color_gradientn(name = "Drought\nresponse\n(logFC)",
                        colors = brewer.pal(11,"BrBG")[11:1],
                        limits = c(-maxval,maxval),
                        values = values,
                        rescaler = function(x, ...) x, oob = identity) +
  scale_alpha_discrete(range = c(0, 1), guide = FALSE) +
  ggtitle("Davis Rhizosphere") +
  ylab("Normalized Counts") +
  #scale_y_log10() +
  theme_light() +
  scale_fill_manual(values = c("#D8B365" , "#5AB4AC")) +
  theme(text = element_text(size = 15))

otu.b <- otu.vsd.tidy %>% 
  inner_join(clt.fc.sig, by = c("Compartment", "Soil", "Cultivar", "OTU_ID")) %>% 
  filter(Soil == "B") %>%
  ggplot(aes(Cultivar, RelAb, fill = Treatment)) +
  geom_boxplot() +
  geom_point(aes(Cultivar, -0.5, color = estimate, alpha = p.adjusted < 0.05), size = 5) +
  facet_wrap(~OTU_ID, scales = "free_y", ncol = 6) +
  scale_color_gradientn(name = "Drought\nresponse\n(logFC)",
                        colors = brewer.pal(11,"BrBG")[11:1],
                        limits = c(-maxval,maxval),
                        values = values,
                        rescaler = function(x, ...) x, oob = identity) +
  scale_alpha_discrete(range = c(0, 1), guide = FALSE) +
  ggtitle("Biggs Rhizosphere") +
  ylab("Normalized Counts") +
  #scale_y_log10() +
  theme_light() +
  scale_fill_manual(values = c("#D8B365" , "#5AB4AC")) +
  theme(text = element_text(size = 15))
  
otu.d; otu.b

```

