---
Phylum-level analysis
---
This script requires the files generated in the following scripts
OTU_filtering.Rmd
drght_contrasts.Rmd
main_effects_phylum.Rmd


Load libraries
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(RColorBrewer)
library(scales)
```

Load data
```{r}
phycla <- readRDS("../Data/phycla_all.RDS")
tax <- readRDS("../Data/tax.RDS")
map <- readRDS("../Data/map.RDS")
```

Relevel and relabel some of the factors in the mapping file
```{r}
###Since soil only has 4 samples per treatment, we need to have a special label for plotting. This generates dummy variables that will be used in the facets

map$Treatment2 <- factor(map$Treatment,
                         levels = c("WC", "DS"),
                         labels = c("C", "D"))
map$Treatment <- factor(map$Treatment,
                        levels = c("WC", "DS"),
                        labels = c("Control", "Drought"))

map$Treatment2 <- ifelse(map$Compartment == "BS", as.character(map$Treatment2), as.character(map$Treatment))

map$Compartment <- factor(map$Compartment,
                          levels = c("BS","RS","ES"),
                          labels = c("BS", "Rhizosphere", "Endosphere"))
map$Soil <- factor(map$Soil,
                   levels = c("A", "B", "D"),
                   labels = c("Arbuckle", "Biggs", "Davis"))

```


This code generates the phylum profiles
```{r}
# Get the relative abundances and reformat
phy.ra <- t(t(phycla)/colSums(phycla))

phy.ra.tidy <- data.frame(phy.ra) %>%
  mutate(Phylum = row.names(phy.ra)) %>%
  gather(key = "SampleID", value = "RelAb", -Phylum)

# Get the top taxa in the whole dataset
phy.ord <- data.frame(Phylum = row.names(phy.ra), Total = rowSums(phy.ra)/sum(phy.ra)) %>%
  arrange(desc(Total)) %>%
  mutate(CumSum = cumsum(Total))

phy.top <- c(as.character(head(phy.ord, n = 12)$Phylum))
  
# Generate the df that will be used for plotting
prof.plot <- phy.ra.tidy %>%
  filter(Phylum %in% phy.ord$Phylum[1:12]) %>%
  inner_join(map, by = "SampleID")
  
# Reorganize the taxa so that all Proteobacteria are at the end, and generate new labels to use greek letters
prof.plot$Phylum <- factor(prof.plot$Phylum,
                           levels = c("Acidobacteria",
                "Actinobacteria",
                "Bacteroidetes",
                "Chloroflexi",
                "Firmicutes",
                "Gemmatimonadetes",
                "Planctomycetes",
                "Verrucomicrobia",
                "Proteobacteria.Alphaproteobacteria",
                "Proteobacteria.Betaproteobacteria",
                "Proteobacteria.Deltaproteobacteria",
                "Proteobacteria.Gammaproteobacteria"))

my.labs <- list("Acidobacteria",
                "Actinobacteria",
                "Bacteroidetes",
                "Chloroflexi",
                "Firmicutes",
                "Gemmatimonadetes",
                "Planctomycetes",
                "Verrucomicrobia",
                expression(paste(alpha, " - Proteobacteria")),
                expression(paste(beta, " - Proteobacteria")),
                expression(paste(delta, " - Proteobacteria")),
                expression(paste(gamma, " - Proteobacteria")))

# Plot
phyloplot <- ggplot() +
  geom_bar(data = prof.plot, aes(paste(Cultivar, Tub), RelAb, fill = Phylum), stat = "identity", position = "stack") +
  facet_grid(Soil ~ Compartment + Treatment2, scales = "free", space = "free") +
  scale_fill_manual(name = "Phylum",
                    values = c(brewer.pal(8, "Set2")[1:8], brewer.pal(11, "RdYlBu")[7:10]),
                    labels = my.labs) +
  geom_point(data = filter(prof.plot, Soil == "Davis" & Compartment != "BS"), 
             aes(paste(Cultivar, Tub), -0.08, color = Cultivar), 
             size = 5, shape = 15) +
  scale_color_manual(values = c(brewer.pal(12, "Paired")[7:10]),
                     guide = FALSE) +
  labs(y = "Relative Abundance") +
  theme_light() +
  theme(text = element_text(size = 17),
        legend.position = "right",
        legend.text = element_text(size = 16),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  guides(fill = guide_legend(ncol = 1))

phyloplot
  
```

This takes a look at the main effects on the relative abundances of phyla
```{r}

effects <- readRDS("../Tables/main_eff_phy.RDS")

# Get the stats  
main.eff.stats <- effects %>% 
  filter(OTU_ID != "unassigned" & OTU_ID != "unclassified" & OTU_ID != "Proteobacteria.unclassified") %>%
  filter(p.adjusted < 0.05) %>% 
  inner_join(phy.ord, by = c("OTU_ID" = "Phylum")) %>%
  group_by(Effect) %>% 
  summarise(n = sum(estimate != 0),
            CumRelAb = sum(Total))

# Generate the data frame that will be used for plotting
plot.effect <-effects %>%
  filter(p.adjusted < 0.05) %>%
  filter(OTU_ID != "unassigned" & OTU_ID != "unclassified" & OTU_ID != "Proteobacteria.unclassified") %>%
  select(OTU_ID) %>%
  inner_join(effects, by = "OTU_ID") %>%
  inner_join(phy.ord, by = c("OTU_ID" = "Phylum"))

plot.effect$OTU_ID <- gsub(x = plot.effect$OTU_ID, pattern = "Proteobacteria.", replacement = "") 

# Plot the log p.adjusted value of the significant taxa per main effect
main.eff.logp <- plot.effect %>%
  ggplot(aes(reorder(OTU_ID, -Total), Effect, fill = -log10(p.adjusted))) +
  geom_tile(aes(alpha = p.adjusted < 0.05), color = "gray", size = 1) +
  scale_fill_gradientn(name = "-log10(P)",
                       colors = RColorBrewer::brewer.pal(9,"BuPu"), trans = "sqrt") +
  scale_alpha_discrete(range = c(0, 1), guide = FALSE) +
  geom_point(data = filter(plot.effect, Effect == "Compartment"), 
             aes(reorder(OTU_ID, -Total), -0.2, color = Total),
             size = 8, shape = 15) +
  scale_color_gradientn(name = "Relative Abundance",
                       colors = RColorBrewer::brewer.pal(9,"Greys")[3:9],
                       trans = "sqrt") +
  theme_light()+
  theme(text = element_text(size = 15),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 0),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"),
        panel.spacing = unit(0.5, "lines"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        legend.position = "bottom") +
  guides(fill = guide_colorbar(title.position = "top", 
                               title.hjust = 0.5,
                               barwidth = 20,
                               barheight = 1,
                               reverse = TRUE),
         color = guide_colorbar(title.position = "top", 
                               title.hjust = 0.5,
                               barwidth = 20,
                               barheight = 1,
                               reverse = TRUE))

# Plot the cumulative relative abundance of the significant taxa per main effect
main.eff.cra <- effects %>% 
  filter(OTU_ID != "unassigned" & OTU_ID != "unclassified" & OTU_ID != "Proteobacteria.unclassified") %>%
  filter(p.adjusted < 0.05) %>% 
  inner_join(phy.ord, by = c("OTU_ID" = "Phylum")) %>%
  group_by(Effect) %>% 
  summarise(C.R.A. = sum(Total)) %>%
  ggplot(aes(Effect, C.R.A.)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_light()+
  theme(text = element_text(size = 15),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"))
  

main.eff.stats; main.eff.logp ; main.eff.cra
```

Generate the logFC plot for drought responsive taxa in each soil and compartment
```{r}

paired.wt <- readRDS("../Tables/dab_drought.RDS")
paired.wt <- filter(paired.wt, Compartment != "WC")

all.wt <- paired.wt %>%
  filter(Rank == "PhyClass") %>%
  filter(p.adjusted < 0.05) %>%
  filter(OTU_ID != "unassigned" & OTU_ID != "unclassified" & OTU_ID != "Proteobacteria.unclassified") %>%
  select(OTU_ID) %>%
  inner_join(paired.wt, by = "OTU_ID") %>%
  filter(Rank == "PhyClass") %>%
  inner_join(phy.ord, by = c("OTU_ID" = "Phylum"))

all.wt$Compartment <- factor(all.wt$Compartment,
                        levels = c("BS","RS", "ES"),
                        labels = c("Bulk Soil", "Rhizosphere", "Endosphere"))

all.wt$Soil <- factor(all.wt$Soil,
                      levels = c("A", "B", "D"),
                      labels = c("Arbuckle", "Biggs", "Davis"))

all.wt$OTU_ID <- gsub(x = all.wt$OTU_ID, pattern = "Proteobacteria.", replacement = "") 


maxval  <- ceiling(max(abs(all.wt$estimate)))
values  <- seq(-maxval,maxval,length = 8)

all.wt$OTU_ID <- factor(all.wt$OTU_ID)

fc.plot <- all.wt %>%
  ggplot(aes(reorder(OTU_ID, - Total), Soil, fill = estimate)) +
  geom_tile(aes(alpha = p.adjusted < 0.05), color = "gray", size = 1) +
  scale_fill_gradientn(name = "logFC",
                        colors = brewer.pal(11,"BrBG")[11:1],
                        limits = c(-maxval,maxval),
                        values = values,
                        rescaler = function(x, ...) x, oob = identity) +
  scale_alpha_discrete(range = c(0, 1), guide = FALSE) +
  facet_grid(Compartment ~ ., scales = "free", space = "free") +
  theme_light()+
  theme(text = element_text(size = 15),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 0),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"),
        panel.spacing = unit(0.5, "lines"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        legend.position = "right") +
  guides(fill = guide_colorbar(title.position = "top", 
                               title.hjust = 0.5,
                               barwidth = 1.5,
                               barheight = 7))

fc.plot
```
